Machine Learning based Translation Initiation Rate predictor

### 2020-11-24 paper/main.tex
Small changes.

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}. In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device. Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT. Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:

@@

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red. The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
    \label{fig: Anatomy of the randomized sequence.}
\end{figure}

@@

For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly.
The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
Figure X shows the results of the zeroth and all the subsequent rounds.\\

>>>>

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}. In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red. The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
    \label{fig:Anatomy}
\end{figure}

In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device. Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT. Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:

@@

@@

For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly.
The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
Figure X shows the results of the zeroth and all the subsequent rounds.\\

### 2020-11-24 paper/methods.tex
create paper folder
Small changes.

\textbf{Plasmid Design.}
We have used the pBbB6c-GFP plasmid for all our designs.
This plasmid comes with GFP mut3b CDS inducible with addition of IPTG.
The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
Primers and the assembly strategy have been generated using the Teselagen DESIGN software (Teselagen Biotechnology).

@@

@@

\begin{figure}[t]
\centering
\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
\end{minipage}
\caption{Batch Recommendation. We use the batch size of 2, with 5 initial observations. The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e. {-2, -1.5, -1, ..., 9.5, 10}.
The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
(a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
(b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
\label{fig:batch rec}
\end{figure}

>>>>

\textbf{Plasmid Design.}
We have used the pBbB6c-GFP plasmid for all our designs.
This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl β-D-1-thiogalactopyranoside (IPTG).
The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
Primers and the assembly strategy have been generated using the Teselagen DESIGN software (Teselagen Biotechnology).

@@

\textbf{Sequencing}
Sequences that showed TIR on the level of at least 85\% of TIR of the consensus sequence were sequenced using Sanger sequencing either at the Australian Genome Research Facility (Brisbane, QLD, Australia) or at Macrogen (Seoul, Republic of Korea).

@@

\begin{figure}[t]
\centering
\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}}\par\medskip
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
\end{minipage}
\caption{Batch Recommendation. We use the batch size of 2, with 5 initial observations. The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e. {-2, -1.5, -1, ..., 9.5, 10}.
The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
(a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
(b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
\label{fig:batch rec}
\end{figure}
